Adapting the MVDR to a GPU for active sonar systems of different sizes.

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	04:11:31	(Wed23.Jan)

We propose three different solutions to the complexity issue depending on system size. For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU. For medium sized systems we suggest applying the MVDR in the frequency domain (beamspace) instead, which allows for comparable processing rates at a negligible reduction in image quality. Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
% 
% 
% The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
% \\
% << MOTIVATION >>
% \\

Comparable work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\

@@

A few properties let us improve these matters notably. First, we may infer from (\ref{spatialR}) that:
\begin{itemize}
\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one triagonal in them.
\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
\end{itemize}
Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.

@@

As we increase system size the inversion step will become more of a bottleneck. Although the dimensionality of the covariance matrix is reduced to $L$ due to subarray averaging, we can reduce it further by performing MVDR beamforming. The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array angles of arrival in the spatial domain; the zero-frequency correspond to broadside, and increasing frequencies map to monotonously increasing opening angles. Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where the signal came from.

@@

Our GPU implementation consists of three main kernels relating to the computing steps illustrated in Figure \ref{mvdr_implementation}. In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$. All pixels are processed before moving on to the next step. Our efforts here were mainly focused on estimating $\eR$, because an official batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.

>>>>

We propose three different solutions to the complexity issue depending on system size. For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU. For medium sized systems we suggest applying the MVDR in the beamspace (frequency domain) instead, which allows for comparable processing rates at a negligible reduction in image quality. Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
% 
% 
% The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
% \\
% << MOTIVATION >>
% \\

Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\

@@

A few properties let us improve these matters notably. First, we may infer from (\ref{spatialR}) that:
\begin{itemize}
\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
\end{itemize}
Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.

@@

As we increase system size the inversion step will take over as the main bottleneck. We can, however, reduce the dimensionality further than the subarray size of $L$ by performing MVDR beamforming the frequency domain. The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array and angles of arrival in the spatial domain; the zero-frequency will correspond to broadside, and increasing frequencies will map to monotonously increasing opening angles. Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where our signals are coming from~\cite{Nilsen2009a,VanTrees2002}. The method is named beamspace MVDR.

@@

Our GPU implementation consists of three main kernels relating to the computing steps illustrated in Figure \ref{mvdr_implementation}. In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$. All pixels are processed by one step before we move on to the next step. Our efforts here were mainly focused on estimating $\eR$, because an official batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
